### 2018-05-16 submission/manuscript.Rmd
re-write abstract

**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow. Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and polymerase can change the bacterial community and introduce bias to the final obtained results.

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. The analysis initially examined the number of Operational Taxonomic Units (OTUs) for both fecal samples and mock communities. It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, and chimera prevalence. Our analysis also examined chimera prevalence correlation with the number of OTUs.

**Results.** When analyzing fecal samples we observed a different number of OTUs between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Our analysis identified these polymerase dependent differences in the number of OTUs as early as 20 cycles in the mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by  polymerase and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs which was not affected by chimera removal with VSEARCH.

>>>>

**Background.** Bias is introduced at different stages of the 16S rRNA gene sequencing workflow. Although the number of cycles and high fidelity (HiFi) DNA polymerase are studied less often, they are still important sources of bias in this workflow. Here, we examine how both cycle number and polymerase can change the bacterial community and introduce both error and bias to the final obtained results.

**Methods.** We extracted DNA from fecal samples (n=4) using a PowerMag DNA extraction kit with a 10 minute bead beating step and amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa, Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock communities (technical replicates n=4) consisting of previously isolated whole genomes of 8 different bacteria used the same approach. The analysis initially examined the number of Operational Taxonomic Units (OTUs) for fecal samples and mock communities. It also assessed polymerase dependent differences in the Bray-Curtis index, error rate, sequence error prevalence, chimera prevalence, and chimera prevalence correlation with the number of OTUs.

**Results.** When analyzing fecal samples we observed a different number of OTUs between HiFi DNA polymerases at 35 cycles (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). Our analysis identified these polymerase dependent differences in the number of OTUs as early as 20 cycles in mock communities (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). Chimera prevalence varied by  polymerase and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs.

### 2018-05-16 submission/manuscript.tex
re-write abstract

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Although the number of cycles and high
fidelity (HiFi) DNA polymerase are studied less often, they are still
important sources of bias in this workflow. Here, we examine how both
cycle number and polymerase can change the bacterial community and
introduce bias to the final obtained results.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for both fecal samples and mock communities. It also assessed
polymerase dependent differences in the Bray-Curtis index, error rate,
sequence error prevalence, and chimera prevalence. Our analysis also
examined chimera prevalence correlation with the number of OTUs.

\textbf{Results.} When analyzing fecal samples we observed a different
number of OTUs between HiFi DNA polymerases at 35 cycles (P-value
\textless{} 0.0001). Our analysis identified these polymerase dependent
differences in the number of OTUs as early as 20 cycles in the mock
communities (P-value = 0.002). Chimera prevalence varied by polymerase
and this variation persisted after chimera removal using VSEARCH. We
also observed positive correlations between chimera prevalence and the
number of OTUs which was not affected by chimera removal with VSEARCH.

>>>>

\textbf{Background.} Bias is introduced at different stages of the 16S
rRNA gene sequencing workflow. Although the number of cycles and high
fidelity (HiFi) DNA polymerase are studied less often, they are still
important sources of bias in this workflow. Here, we examine how both
cycle number and polymerase can change the bacterial community and
introduce both error and bias to the final obtained results.

\textbf{Methods.} We extracted DNA from fecal samples (n=4) using a
PowerMag DNA extraction kit with a 10 minute bead beating step and
amplified at 15, 20, 25, 30, and 35 cycles using Accuprime, Kappa,
Phusion, Platinum, or Q5 HiFi DNA polymerase. Amplification of mock
communities (technical replicates n=4) consisting of previously isolated
whole genomes of 8 different bacteria used the same approach. The
analysis initially examined the number of Operational Taxonomic Units
(OTUs) for fecal samples and mock communities. It also assessed
polymerase dependent differences in the Bray-Curtis index, error rate,
sequence error prevalence, chimera prevalence, and chimera prevalence
correlation with the number of OTUs.

\textbf{Results.} When analyzing fecal samples we observed a different
number of OTUs between HiFi DNA polymerases at 35 cycles (P-value
\textless{} 0.0001). Our analysis identified these polymerase dependent
differences in the number of OTUs as early as 20 cycles in mock
communities (P-value = 0.002). Chimera prevalence varied by polymerase
and this variation persisted after chimera removal using VSEARCH. We
also observed positive correlations between chimera prevalence and the
number of OTUs.
